---
title: "One-sample expansion cohort"
author: "Emily C. Zabor"
date: "Last updated: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{One-sample expansion cohort}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Introduction

The goal of this article is to provide a tutorial on the use of the `calc_posterior()` and `calc_predictive()` functions in the context of a one-sample expansion cohort. It is becoming increasingly common for expansion cohorts to be included at the end of phase I trials. An expansion cohort is a group of additional patients being treated at the recommended phase 2 dose (RP2D) that was identified in the dose finding portion of the phase I study, and is typically used to further characterize the toxicity and/or efficacy of the RP2D. Since one goal of expansion cohorts is to conserve the limited available resources for clinical studies, early stopping for futility is an important consideration. In this article we demonstrate the use of sequential posterior probability monitoring for futility in the context of a one-sample expansion cohort.


# Setup

To execute the code in this tutorial, you first need to install and load the {ppseq} package.

```{r setup}
# remotes::install_github("zabore/ppseq")
library(ppseq)
```


# Methods

We want to plan a one-sample study. The unacceptable response rate is 0.1, and the acceptable response rate 0.3. We want to look after every 5 patients, with a maximum sample size of 50.

Let's start by seeing what happens in a single simulated trial. We use the function `sim_single_trial()` to simulate response counts based on the properties of our trial and calculate the posterior predictive value of exceeding the target posterior probability of success.

Let's say, for simplicity, that we would stop for futility if our posterior predictive probability of exceeding the target posterior probability, which in this case is 0.95, falls below 0.1. 

First let's consider a trial where we achieve our acceptable response rate of 0.3.

```{r}
set.seed(123)

sim_single_trial(prob = 0.3, n = seq(5, 50, 5), direction = "greater", p = 0.1,
                 delta = NULL, prior = c(0.5,0.5), S = 5000, 
                 N = 50, theta = 0.95)
```

We see here that the posterior predictive probability, in the column `ppp`, never drops below the futility level of 0.1, so we would continue enrolling in this expansion cohort until the end of the trial, where we see we have reached a 100\% predictive probability of having posterior probability exceeding 0.95. 

Now let's consider a trial where we do not achieve our acceptable response rate of 0.3, but rather observe the unacceptable response rate of 0.1.

```{r}
set.seed(123)

sim_single_trial(prob = 0.1, n = seq(5, 50, 5), direction = "greater", p = 0.1,
                 delta = NULL, prior = c(0.5,0.5), S = 5000, 
                 N = 50, theta = 0.95)
```

We see in this example that we drop below the futility threshold of 0.1 after enrolling 15 patients, at which point we have a posterior predictive probability of 0.0960. This would cause the trial to stop for futility to preserve resources for the investigation of more promising treatments.

But when we are designing a trial, we want to achieve fixed operating characteristics such as type I error and power. There are several goals:

1. Determine the posterior probability threshold that maintains a pre-determined overall type I error rate. For simplicity in the previous examples we used an arbitrary threshold of 0.95, but in practice we would want to calibrate this threshold to the desired design properties of our trial.
2. Determine the predictive probability threshold to stop for futility.
3. Using the identified thresholds, simulate many trials to determine the average probability of stopping early for futility. 
4. Consider different interim monitoring schedules

To accomplish the first goal, we simulate many trials under the null and search over a range of thresholds to identify the threshold that maintains the overall type I error at the desired level.

The `calibrate_posterior_threshold()` function can accomplish this at the final sample size only, not considering any interim looks or early stopping, for the one-sample or two-sample case and any number of thresholds of interest to consider

```{r}
thresholds <- c(0, 0.7, 0.74, 0.78, 0.82, 0.86, 0.9, 0.92, 0.93, 0.94, 0.95,
                0.96, 0.97, 0.98, 0.99, 0.999, 0.9999, 0.99999, 1)

set.seed(123)

res <- calibrate_posterior_threshold(prob = 0.1, N = 50, direction = "greater", 
                              p0 = 0.1, delta = NULL, prior = c(0.5,0.5), 
                              S = 5000, theta = thresholds)

knitr::kable(res)
```

Based on these results, we see that our target type I error falls between a threshold of 0.95 and 0.96. We can refine our target thresholds and recalculate.

```{r}
thresholds <- seq(0.95, 0.96, 0.001)

set.seed(123)

res <- calibrate_posterior_threshold(prob = 0.1, N = 50, direction = "greater", 
                              p0 = 0.1, delta = NULL, prior = c(0.5,0.5), 
                              S = 5000, theta = thresholds)

knitr::kable(res)
```

And we see that a threshold of 0.959 results in a type I error rate of 0.496.

However, what we really want in this case is to simultaneously calibrate the posterior threshold and the predictive threshold, considering the interim looks for futility. The `calibrate_pp_ppp_threshold()` function can accomplish this, and will return the average sample size, the proportion of positive trials, and the proportion of trials that were stopped early for each combination of considered thresholds, thus accomplishing goals 1 through 3 above.

Let's consider the null scenario where the target response probability is 0.1 and the observed response probability is also 0.1. We generate 1000 simulated datasets. Note that while this function does operate in parallel using functions from the `future` and `furrr` packages, it still is quite time consuming. Just a single simulated trial takes about 2 minutes to run. As a result, this code is not evaluated here, but was rather run previously and the saved results are presented.

```{r eval = FALSE}
set.seed(123)

tic()
type1_tbl <- calibrate_pp_ppp_threshold(prob = 0.1, n = seq(5, 25, 5),
                                        direction = "greater", p0 = 0.1,
                                        delta = NULL, prior = c(0.5, 0.5),
                                        S = 5000, N = 25, nsim = 1000,
                                        pp_threshold = c(0.9, 0.95, 0.96, 0.98),
                                        ppp_threshold = seq(0.05, 0.2, 0.05))
toc()

save(type1, file = here::here("vignettes", "type1_tbl.rda"))
```

```{r eval = FALSE}
load(here::here("vignettes", "type1_tbl.rda"))

knitr::kable(type1)
```