---
title: "One-sample expansion cohort"
author: "Emily C. Zabor"
date: "Last updated: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{One-sample expansion cohort}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Introduction

The goal of this article is to provide a tutorial on the use of the `calc_posterior()` and `calc_predictive()` functions in the context of a one-sample expansion cohort. It is becoming increasingly common for expansion cohorts to be included at the end of phase I trials. An expansion cohort is a group of additional patients being treated at the recommended phase 2 dose (RP2D) that was identified in the dose finding portion of the phase I study, and is typically used to further characterize the toxicity and/or efficacy of the RP2D. Since one goal of expansion cohorts is to conserve the limited available resources for clinical studies, early stopping for futility is an important consideration. In this article we demonstrate the use of sequential posterior probability monitoring for futility in the context of a one-sample expansion cohort.


# Setup

To execute the code in this tutorial, you first need to install and load the {ppseq} package.

```{r setup}
# remotes::install_github("zabore/ppseq")
library(ppseq)
```


# Methods

We want to plan a one-sample study. The unacceptable response rate is 0.1, and the acceptable response rate 0.3. We want to look after every 5 patients, with a maximum sample size of 50.

Let's start by seeing what happens in a single simulated trial. We use the function `sim_single_trial` to simulate response counts based on the properties of our trial and calculate the posterior predictive value of exceeding the target posterior probability of success.

We would stop for futility if our posterior predictive probability of exceeding the target posterior probability, which in this case is 0.95, falls below 0.05. 

(Question: how do people usually determine these futility/efficacy stopping rules? Are they typically calibrated in some way, or just "arbitrary"?)

First let's consider a trial where we achieve our acceptable response rate of 0.3.

```{r}
sim_single_trial(prob = 0.3, n = seq(5, 50, 5), direction = "greater", p = 0.1,
                 delta = NULL, prior = c(0.5,0.5), S = 5000, seed = 1, 
                 N = 50, theta = 0.95)
```

We see here that the posterior predictive probability, in the column `pp`, never drops below the futility level of 0.05, so we would continue enrolling in this expansion cohort until the end of the trial, where we see we have reached a 100\% predictive probability of having posterior probability exceeding 0.95. 

Now let's consider a trial where we do not achieve our acceptable response rate of 0.3, but rather observe the unacceptable response rate of 0.1.

```{r}
sim_single_trial(prob = 0.1, n = seq(5, 50, 5), direction = "greater", p = 0.1,
                 delta = NULL, prior = c(0.5,0.5), S = 5000, seed = 1, 
                 N = 50, theta = 0.95)
```

We see in this example that we drop below the futility threshold of 0.05 after enrolling just 10 patients, at which point we have a posterior predictive probability of 0.0390. This would cause the trial to stop for futility to preserve resources for the investigation of more promising treatments.

But when we are designing a trial, what we want is to know the operating characteristics across many simulated trials. There are two goals:

1. Determine the posterior probability threshold that maintains a pre-determined overall type I error rate. For simplicity in the previous examples we used an arbitrary threshold of 0.95, but in practice we would want to calibrate this threshold to the desired design properties of our trial.
2. Once the posterior probability threshold is established, we want to simulate many trials to determine the average probability of stopping early for futility. 

To accomplish the first goal, we simulate many trials under the null and searching over a range of thresholds to identify the threshold that maintains the overall type I error at the desired level.

(Note to self: since in this example we only have futility monitoring, we only have to consider rejection of the null hypothesis at the end of the trial. So if at the end of the trial P(p1 > p0) exceeds the threshold under consideration, that would be a type I error)

To accomplish the second goal, we simulate many trials under the alternative and using the identified threshold and we average the posterior predictive probabilities to assess the probability of stopping for futility.

We generate response counts based on the null response rate of 0.1 for a sample size of 50 for 5000 independent trials. For each resulting response count, we then calculate the posterior probability Pr(p1 > p0) using the function `calc_posterior()`. We compare the resulting posterior probabilities to the thresholds of interest and calculate the type I error rate as the average number of trials in which the posterior probability exceeded the threshhold. 

(Note to self: turn the below into a function)

```{r}
library(dplyr)

# Note: we don't want to set a seed within the calc_posterior() function in this case as that would return the same results every time and the goal is to return a variety of results, so we want to set an overall seed instead
set.seed(20210310) 

posts <-
  purrr::map_dbl(1:5000, 
          ~stats::rbinom(n = 1, size = 50, prob = 0.1) %>% 
            purrr::map_dbl(
              ~calc_posterior(y = .x, n = 50, direction = "greater", p0 = 0.1,
                              delta = NULL, prior = c(0.5,0.5), S = 5000, 
                              seed = NULL)
              )
          )

typeIerr_tbl <-
  tibble::tibble(
    theta = c(0, 0.7, 0.74, 0.78, 0.82, 0.86, 0.9, 0.92, 0.93, 0.94, 0.95, 
              0.96, 0.97, 0.98, 0.99, 0.999, 0.9999, 0.99999, 1), 
    typeIerr = c(0, 0.7, 0.74, 0.78, 0.82, 0.86, 0.9, 0.92, 0.93, 0.94, 0.95, 
                 0.96, 0.97, 0.98, 0.99, 0.999, 0.9999, 0.99999, 1) %>% 
      purrr::map_dbl(~mean(posts > .x))
    )

knitr::kable(typeIerr_tbl, digits = c(5, 3))
```

Based on these results, we would select a threshold of 0.96 to control the type I error at 0.048. 

(Question: do people usually refine this more to try to directly target 0.05? There could be a second grid search between 0.95 and 0.96 to refine it if so)

Now that we have our target threshold, we will identify the probability of stopping early for futility averaged over many trials.

```{r eval = FALSE}
set.seed(20210309)
pps <-
  purrr::map(
    1:10,
    ~sim_single_trial(prob = 0.3, n = seq(5, 50, 5), direction = "greater", 
                      p = 0.1, delta = NULL, prior = c(0.5,0.5), S = 500, 
                      seed = NULL, N = 50, theta = 0.96)
    )
```


```{r}
knitr::knit_exit()
```




```{r eval = FALSE}
# This is just me trying to figure out Brian's code from the Covid trial

 # prob y1 < y0
 posterior_y0.gt.y1 <- function(n0, n1, y0, y1, delta=0, prior=c(0.5,0.5)){ ## y0 standard     y1 experimental
  SAMP <- 10000
  out <- sum( rbeta(SAMP, prior[1] + y0, prior[2] + n0 - y0) > ( rbeta(SAMP, prior[1] + y1, prior[2] + n1 - y1) + delta ) ) / SAMP
  return(out)
 }

find.Boundary <- function(j, theta, Tab, delta=0, prior=c(0.5,0.5) ){

 Y=Tab$n1[j]; STOP <- FALSE
 temp <- posterior_y0.gt.y1( n0=Tab$n0[j], n1=Tab$n1[j], y0=Tab$y0[j], y1=Y, delta=delta, prior=prior )
 if( is.na(temp) ){ temp <- 0; print(paste0("NA - ",Tab[j,])) }
 if( temp > theta ){ STOP <- TRUE }
 while( !STOP ){
  Y <- Y - 1
  temp <- posterior_y0.gt.y1( n0=Tab$n0[j], n1=Tab$n1[j], y0=Tab$y0[j], y1=Y, delta=delta, prior=prior )
  if( is.na(temp) ){ temp <- 0 }
  if( Y == 0 ){ STOP <- TRUE }
  if( temp > theta ){ STOP <- TRUE }
 }

 if( temp > theta ){ out <- Y 
 } else{ out <- NA }

 return( c(j = j, out = out) )
}

Max <- 2000
N.tab <- expand.grid( list( seq(100,1500,100), seq(100,1500,100) ) )
N.tab <- N.tab[-which( rowSums(N.tab) > Max ), ]

Tab <- as.data.frame( list(n0=NA, n1=NA, y0=NA) ) 
for(i in 1:nrow(N.tab)){

 s <- 0:N.tab[i,1]
 temp <- as.data.frame( list(n0=rep(N.tab[i,1],length(s)),
                             n1=rep(N.tab[i,2],length(s)),
                             y0=s) )
 Tab <- rbind(Tab, temp)
}
Tab <- Tab[-1,]
 

 prior <- c(0.1,0.9)
 delta <- 0
 theta <- 0.9516

Tab2 <- Tab[1:101, ]

temp <- 
  1:nrow(Tab2) %>% 
  map_df(~find.Boundary(j = .x, theta = 0.9516, Tab = Tab2, delta = 0, 
                     prior = c(0.1,0.9)))

 ### Merge Boundaries ###
 Tab2$y1.stop <- rep(NA, nrow(Tab2) )
 Tab2$y1.stop[ temp[[1]] ] <- temp[[2]]
```

